Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-07 numerical.tex
mkl edits of 5.3, 6.0, 6.1, 7; atlanta file updates
edits to 6.3.  Also removed max = 3H cases from two figures.  The fist one (beta = 10^-3) is enough to show that it's less convergent.  From there the 5 & 7 cases are clearly the most relevant and we need to narrow the focus onto what's most important.  Minor text changes in 6.2 and fig caption to accommodate this change.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_iso}
  \caption{Pseudo-enthalpy perturbation $W=\delta P/\rho$ (top) and
    vertical velocity perturbation $\delta v_z$ of the fundamental VSI with 
    $\khat=10$ in the fiducial disk with rapid thermal relaxation
    $\beta=10^{-3}$. The real (imaginary) parts of $W$ and $\delta
    v_z$ are plotted as solid (dotted) lines.  
    \label{lowfreq_eigenfunc}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx30_z7_analytic.ps}
  \caption{Same as Fig. \ref{compare_modes_cool_kx10} but for
    $\khat=30$. Examples of surface modes are marked with `s'. 
    \label{compare_modes_cool_kx30} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx30} shows how cooling affects higher wavenumber
modes, specifically $\khat=30$.   We focus on the $\zmax/H = 5, 7$ and analytic results for $\beta \geq 0.03$. 
These cases  show good convergence, though not quite as good as for $\khat = 10$.

@@

Fig. \ref{compare_modes_cool_kx30} also demonstrates that artificial surface modes are damped for $\beta \geq 0.03$.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
Having explored the behavior of VSI modes with cooling, we turn to the numerical vailidity of the 
analytic cooling criterion, $\beta < \beta_\mathrm{crit} = h|q|/(\gamma -1)$.

...

In Fig. \ref{bcrit_compare1} we plot the maximum VSI growth rates
found in the fiducial disk model as a function of the thermal
relaxation timescale $\beta$ for several values of $\khat$ and
$\zmax$. We also plot the critical timescale
$\beta_\mathrm{crit}=0.125$ for the fiducial disk parameters.

Note that the curves are not neccessarily smooth because the most unstable
mode may not always correspond to the same type of disturbance. For example,
as seen above, for $\khat \leq 10$ higher order body modes are
dominant as $\beta\to 0$, but the fundamental 
mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
$\khat=5,\,10$ (and $30$ for $\zmax=3H$), we find
$\beta_\mathrm{crit}$ accurately predicts the thermal timescal beyond
which the VSI is suppressed.

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
$\beta\to0$, we find surface modes dominate so the maximum growth rate
depends on $\zmax$ in the limit of rapid thermal relaxation. 
However, the maximum growth rate is insensitive to 
$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
the stabilization of the fundamental mode.  
%not surprising since we enable strongly stabilizing buoyancy forces
%at large zmax

Our result is consistent with numerical simulations performed by
\citetalias{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
rates for $\beta\gtrsim 0.6$ are small, and are restricted to very
small radial lengthscales $\sim 10^{-2}H$, for which current numerical
simulations may not be able to resolve.
%  To destablize the disk with
% $\beta\gtrsim O(1)$, one might consider 
% $\khat > 100$ to increase the maximum growth rate by enabling surface
% modes, but in a realistic disk
% very small scale disturbances will likely be subject to viscous decay
% (see \S\ref{application}). 
%Numerical simulations are also unlikely to
%resolve such small scales.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z3} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z5} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z7}  
  \caption{Maximum VSI growth rates in the fiducial disk 
     model as a function of the thermal relaxation timescale
     $\beta$, for $\zmax=3H$ (top), $\zmax=5H$ (middle, standard
     setup) and $\zmax=7H$ (bottom). The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

As pointed out by \citetalias{barker15}, in vertically isothermal disks with
an imposed boundary, surface modes are dependent on $\zmax$, but they
are often the most unstable. Despite this caveat, and that surface
modes are not accounted in our analysis, we see from
Fig. \ref{bcrit_compare1} that $\beta_\mathrm{crit}$ nevertheless
provides a reasonable estimate for an upper limit to the thermal
timescale for the VSI.  
%It becomes an accurate criter when considering fixed
%$\khat$

We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$ h\in[0.02,0.1]$ separately. The numerically obtained
$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
result, together with Fig. \ref{bcrit_compare1}, shows that  
$\beta \lesssim \beta_\mathrm{crit}$ is a practical
criteria for the VSI to operate effectively in stably stratified,
vertically isothermal disks.

% The agreement improves with decreasing  $|q|$,
% $ h$ and increasing $\gamma$, i.e. weaker instability. There is
% a noticeable difference as $\gamma\to1$ because the derivation of
% $\beta_\mathrm{crit}$ assumed $\gamma>1$.

@@

>>>>

@@

\begin{figure}
 % \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  %0cm]{figures/compare_modes_cool_kx30_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx30_z7_analytic.ps}
  \caption{Same as Fig. \ref{compare_modes_cool_kx10} but for
    $\khat=30$ and domain sizes $\zmax = 5H$ (\emph{top}) and $7H$ (\emph{bottom}). Examples of surface modes are marked with `s'. 
    \label{compare_modes_cool_kx30} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx30} shows how cooling affects higher wavenumber
modes, specifically $\khat=30$.   We only show the larger domains with $\zmax = 5H, 7H$. 
These cases again show good convergence for $\beta \geq 0.03$, which the
the smaller domain with  $\zmax = 3H$ lacks.

@@

Fig. \ref{compare_modes_cool_kx30} also demonstrates that artificial surface modes are damped for $\beta \geq 0.03$.
We are thus confident that artificial surface modes should not affect our determination of the critical cooling time.

@@

\begin{figure}
  %\includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  %0.9cm]{figures/gcorr_compare_iso_maxrate_z3} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z5} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.9cm]{figures/gcorr_compare_iso_maxrate_z7}  
  \caption{Maximum VSI growth rates in the fiducial disk 
     model as a function of the thermal relaxation timescale
     $\beta$, for %$\zmax=3H$ (top), 
     $\zmax=5H$ (top) and $\zmax=7H$ (bottom). The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

\subsection{Critical thermal relaxation timescale}\label{bcrit_num_test}
Having explored the behavior of VSI modes with cooling, we turn to the numerical validity of the 
analytic cooling criterion for vertically isothermal disks, $\beta < \beta_\mathrm{crit} = h|q|/(\gamma -1)$.

Fig. \ref{bcrit_compare1} shows how VSI growth rates vary with cooling time 
in our fiducial model with $\beta_\mathrm{crit} = 0.125$.  
Curves are for a fixed horizontal wavenumber, and show the maximum growth rate for all vertical mode orders.  The 
discontinuity in some curves occurs when the fastest growth switches to a different mode order.

For $\khat = 5, 10$, the growth rate drops to zero at the expected $\beta = \beta_\mathrm{crit}$.
For longer wavelength modes, with $\khat = 1$, growth persists to slightly longer cooling times.  
This difference is not surprising since our analytic derivation assumed $\khat \gg1$, and 
the change is quantitatively minor.

For shorter wavelengths, with $\khat \geq 30$, growth persists for $\beta$ significantly larger
than $\beta_\mathrm{crit}$.  This tail of growth is partly explained by the breaking of the $\khat < 1/|qh| = 20$ 
approximation, used in  the analytic derivation.  Despite the lack of a clear stability boundary 
at high $\khat$, the $\beta_\mathrm{crit}$ threshold remains useful, since growth rates drop to 
$\lesssim 10\%$ of  their maximum value at  $\beta_\mathrm{crit}$ and continue to fall for larger $\beta$. 
 Moreover, we expect that longer wavelength modes with $\khat \lesssim 20$
are more significant for disk transport, see \S\ref{caveats_visc}.

Comparing the $\zmax/H = 5, 7$ cases in Fig. \ref{bcrit_compare1}, we see that the location of the vertical
boundary has little effect on the critical cooling time.  This agreement occurs despite the fact that peak growth rates
differ as $\beta \to 0$.  We are thus confident that artificial boundary effects, including surface modes, do not
affect our analysis of the critical cooling time.

Our results agree with the vertically isothermal analysis of \citetalias{nelson13} 
which used the same disk parameters as our fiducial model.  The expected $\beta_\mathrm{crit} = 0.125$ 
is consistent with the simulations shown in their Fig.\ 12.  \citetalias{nelson13}  
found nonlinear VSI growth for $\beta = 0.06< \beta_\mathrm{crit}$ but no growth for 
  $\beta = 0.6 > \beta_\mathrm{crit}$.\footnote{Since the dimensionless cooling time $T_{\rm relax}$ in \citetalias{nelson13}
  is normalized to the orbital period, we convert $\beta = \OmK P_{\rm orb} T_{\rm relax} = 2 \pi T_{\rm relax}$.}

@@

Fig.\ \ref{bcrit_compare} confirms that the numerically determined
$\beta_\mathrm{crit}$ shows the expected scaling with disk parameters $h$, $q$, and $\gamma$.
For this test we fix $\khat = 10$ (an appropriate value for all the reasons discussed above) and measure
the smallest $\beta$ value where growth vanishes.  The agreement with the analytic scaling of 
Eq.\ \ref{iso_vsi_cond} is quite good, confirming the applicability of our critical cooling time in
vertically isothermal disks.

%we plot the maximum VSI growth rates
%found in the fiducial disk model as a function of the thermal
%relaxation timescale $\beta$ for several values of $\khat$ and
%$\zmax$. We also plot the critical timescale
%$\beta_\mathrm{crit}=0.125$ for the fiducial disk parameters.   
%
%Note that the curves are not neccessarily smooth because the most unstable
%mode may not always correspond to the same type of disturbance. For example,
%as seen above, for $\khat \leq 10$ higher order body modes are
%dominant as $\beta\to 0$, but the fundamental 
%mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
%can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
%$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
%$\khat=5,\,10$ (and $30$ for $\zmax=3H$), we find
%$\beta_\mathrm{crit}$ accurately predicts the thermal timescal beyond
%which the VSI is suppressed.   
%
%Complications from vertical boundaries arise when considering large
%$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
%0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
%seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
%which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
%$\beta\to0$, we find surface modes dominate so the maximum growth rate
%depends on $\zmax$ in the limit of rapid thermal relaxation. 
%However, the maximum growth rate is insensitive to 
%$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
%the stabilization of the fundamental mode.

%Our result is consistent with numerical simulations performed by
%\citetalias{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
%0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
%rates for $\beta\gtrsim 0.6$ are small, and are restricted to very
%small radial lengthscales $\sim 10^{-2}H$, for which current numerical
%simulations may not be able to resolve.
%
%As pointed out by \citetalias{barker15}, in vertically isothermal disks with
%an imposed boundary, surface modes are dependent on $\zmax$, but they
%are often the most unstable. Despite this caveat, and that surface
%modes are not accounted in our analysis, we see from
%Fig. \ref{bcrit_compare1} that $\beta_\mathrm{crit}$ nevertheless
%provides a reasonable estimate for an upper limit to the thermal
%timescale for the VSI.  
%
%We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
%(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
%point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
%$ h\in[0.02,0.1]$ separately. The numerically obtained
%$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
%comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
%result, together with Fig. \ref{bcrit_compare1}, shows that  
%$\beta \lesssim \beta_\mathrm{crit}$ is a practical
%criteria for the VSI to operate effectively in stably stratified,
%vertically isothermal disks.
